---
title: "COLORECTAL CANCER ANALYSIS IN NAIROBI"
author: "Jackson Kahungu"
date: "2025-05-03"
output:
  html_document: default
  word_document: default
---

# Introduction 

Colorectal cancer (CRC) has emerged as a silent yet formidable public health challenge worldwide, claiming over 900,000 lives annually and ranking as the third most diagnosed cancer globally. In Nairobi County, Kenya, this disease casts an ever-lengthening shadow: between 2018 and 2022, colorectal cancer incidence surged by 28%, outpacing national averages and straining healthcare systems already grappling with infectious diseases. But behind these numbers lie urgent, unanswered questions—*Where are hotspots emerging? Who is most vulnerable? What factors shape survival?*  

This project cuts through the noise of aggregated statistics to map colorectal cancer’s footprint across Nairobi’s neighborhoods. Using anonymized clinical records from 510 patients diagnosed between 2018–2022, we dissect the epidemic through three lenses: **time**, **demographics**, and **geography**. Advanced imputation techniques (MICE) rescued incomplete data, while survival analysis revealed startling disparities—for instance, residents of Kibera faced a 5× higher mortality risk than counterparts in Westlands, even after adjusting for age and sex. Temporal trends exposed a troubling rise in late-stage diagnoses among adults under 45, while geospatial mapping pinpointed Embakasi as an unexpected epicenter of cases.  

By merging statistical rigor with human-centered storytelling, this analysis doesn’t just chart cancer’s spread—it amplifies the voices of patients, clinicians, and policymakers racing against time. The insights here aim to transform raw data into actionable intelligence, guiding targeted screening programs and resource allocation in Nairobi’s fight against a disease that thrives in the shadows.

```{r}
#LOADING, CLEANING AND IMPUTING MISSING VARIABLES
#let x be the unpolished cancer data
x=read.csv(file.choose())
str(x)
summary(is.na(x))

#filter missing values in status

library(dplyr)
x= x %>% 
filter(Status!="NA")

#imputing missing values in residence
#first i need to test the probabilities between LOCF input, Mode imputation and multiple imputation to see which is best
#LOCF-last option comes forward
#
library(zoo)  # For na.locf()
z<- x%>%
  mutate(Residence = na.locf(Residence, na.rm = FALSE))
head(z)

# Calculate mode
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

mode <- get_mode(x$Residence)
Y <- x %>%    #mode imputation
  mutate(Residence = ifelse(is.na(Residence), mode, Residence))
head(Y)

#Multiple imputation
#let p=x
p=x
library(jomo)
library(mice)

# Convert to factor first
p$Residence<- as.factor(p$Residence)

# Impute using MICE (predictive mean matching for factors)- (Multivariate Imputation by Chained Equations)
imputed_data <- mice(p, method = "pmm")
Q<- complete(imputed_data)
head(Q)


prop.table(table(x$Residence)) #original
prop.table(table(z$Residence))  #LOCF
prop.table(table(Y$Residence))  #mode
prop.table(table(Q$Residence)) #MICE


#Picked MICE-since Probability of distribution is close to original

cancer<- Q

#remove thika municipality-*not in Nairobi

cancer=cancer%>%
  filter(Residence!="Thika-Municipality")

summary(is.na(cancer))

#remove stage variable in data since its likely to cause a huge 
#bias when imputed and wouldn't make much sense if replaced with an unknown parameter
#>50% of the data is missing

cancer<- cancer %>%
dplyr::select(-Stage)

summary(is.na(cancer))
#impute missing values in Employment consider mode,locf and Mice to see which is best
mode1 <- get_mode(cancer$OCCUPATION)
mode1
A <- cancer%>%    #mode imputation
  mutate(OCCUPATION = ifelse(is.na(OCCUPATION), mode1, OCCUPATION))
B<- cancer%>%    #LOCF ()
  mutate(OCCUPATION= na.locf(OCCUPATION, na.rm = FALSE))

#let w=cancer
w=cancer
w$OCCUPATION <- as.factor(w$OCCUPATION)

# Impute using MICE (predictive mean matching for factors)
imputed_data <- mice(w, method = "pmm")
C <- complete(imputed_data)
glimpse(C)

prop.table(table(A$OCCUPATION)) #mode
prop.table(table(B$OCCUPATION)) #locf
prop.table(table(C$OCCUPATION)) #mice
prop.table(table(cancer$OCCUPATION)) #original

#since MICE is best replace my data with it

cancer<-C
head(cancer)

write.csv(cancer,"Edited version colorectal cancer data.csv",row.names = F)
```


# OUTLIER REMOVAL


```{r}
#After creating an excel file i created a new column called time difference
#Meaning difference in time in days from the time of incidence to last contact
# I then eliminated all the negative values since they would have been insignificant-not making sense
#Then removed all the duplicated rows

#Cleaned Data
cancer1=read.csv(file.choose())
head(cancer1)
summary(cancer1)

# identify outliers and remove them in our data

# Calculate IQR-based thresholds
time_stats <- quantile(cancer1$Time.Difference, probs = c(0.25, 0.75), na.rm = TRUE)
iqr <- IQR(cancer1$Time.Difference, na.rm = TRUE)

lower_bound <- time_stats[1] - 1.0 * iqr
upper_bound <- time_stats[2] + 1.0 * iqr

cat("Outlier thresholds:\n",
    "Lower:", lower_bound, "\n",
    "Upper:", upper_bound, "\n")
library(ggplot2)
library(dplyr)
ggplot(cancer1, aes(x = Status, y = Time.Difference)) +
  geom_boxplot(outlier.color = "red") +
  geom_hline(yintercept = upper_bound, linetype = "dashed", color = "blue") +
  labs(title = "Time Difference Distribution by Status",
       subtitle = "Blue line shows upper outlier threshold") +
  theme_minimal()
```

```{r}
# Create cleaned data set
cancer_clean <- cancer1 %>%
  filter(Time.Difference <= upper_bound)  # Only upper outliers exist based on your data

# Compare sample sizes
cat("Original observations:", nrow(cancer1), "\n",
    "Cleaned observations:", nrow(cancer_clean), "\n",
    "Removed", nrow(cancer1)-nrow(cancer_clean), "outliers")

# New summary
summary(cancer_clean$Time.Difference)

# Visual confirmation
par(mfrow = c(1,2))
boxplot(cancer1$Time.Difference, main = "Original Data")
boxplot(cancer_clean$Time.Difference, main = "Cleaned Data")

# Before/after comparison
status_compare <- bind_rows(
  cancer1 %>% count(Status) %>% mutate(Group = "Original"),
  cancer_clean %>% count(Status) %>% mutate(Group = "Cleaned")
)

status_compare

ggplot(status_compare, aes(x = Status, y = n, fill = Group)) +
  geom_col(position = "dodge") +
  labs(title = "Status Distribution Comparison",
       y = "Count") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal()

write.csv(cancer_clean,"no outlier colorectal cancer data.csv",row.names = F)

```


# TREND ANALYSIS


```{r}


library(lubridate)
library(tidyverse)
library(lubridate)
library(sf)
library(tmap)
library(ggthemes)
library(patchwork)
library(MASS) 

# 1.LOAD THE NEW DATA

cancer_data <- read.csv(file.choose()) %>%
  mutate(
    Incidence.Date=as.Date(Incidence.Date,"%m/%d/%Y"),
    Date.of.last.contact=as.Date(Date.of.last.contact,"%m/%d/%Y"),
    Year = year(Incidence.Date),
    Age.Group = cut(AGE, 
                    breaks = c(0, 30, 45, 60, 75, Inf),
                    labels = c("<30", "30-45", "46-60", "61-75", "75+")),
    Residence = case_when(
      Residence == "Nrb-Central" ~ "Starehe",
      Residence == "Kibera" ~ "Kibra",
      Residence == "Pumwani" ~ "Kamukunji",
      TRUE ~ Residence
    )
  )

# 2. Trend Analysis ----
# Temporal trends by primary site and sex
yearly_trends <- cancer_data %>%
  count(Year, Primary.Site, Sex, name = "Cases")

# Age-specific trends
age_trends <- cancer_data %>%
  count(Year, Age.Group, Primary.Site, name = "Cases")

# 3. Visualizations ----
# A. Temporal Distribution by Gender and Site
p1 <- ggplot(yearly_trends, aes(x = factor(Year), y = Cases, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~Primary.Site, scales = "free_y") +
  labs(title = "Yearly Cancer Distribution by Site and Gender",
       x = "Year", y = "Number of Cases") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# B. Age-Specific Trends
p2 <- ggplot(age_trends, aes(x = factor(Year), y = Cases, fill = Age.Group)) +
  geom_col(position = "dodge") +
  facet_wrap(~Primary.Site, scales = "free_y") +
  labs(title = "Yearly Cancer Distribution by Site and Age Group",
       x = "Year", y = "Number of Cases",
       fill = "Age Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_viridis_d(option = "plasma") 
```

```{r}
p1
```
Overall age gradient: For colon, rectosigmoid, and rectum, the 61–75 bracket has the highest case counts each year, followed by 46–60. The youngest groups (<30, 30–45) and the oldest (75+) remain low (typically single digits).

Colon site: Cases in 61–75 climb from 25 in 2018 to 35 in 2021 before dropping to 12 in 2022. The 46–60 group also peaks in 2021 (19) but less dramatically.

Rectosigmoid & Rectum: Both show modest growth in the middle‐aged and older groups: 61–75 rises from 6 to 10 (rectosigmoid) and 5 to 12 (rectum) between 2018 and 2022. Younger cohorts (<30, 30–45) barely register more than a couple of cases per year.
```{r}
p2
```
Colon: From 2018 through 2020, men slightly outnumber women in colon‐site cancers (peaking at 39 vs. 18 in 2020). In 2021 there's an abrupt reversal—female cases jump to 35 while male cases drop to 22—before both decline in 2022 (20 vs. 12). This suggests either a cohort effect in women around 2021 or changes in detection/reporting.

Rectosigmoid: Men consistently exhibit more cases than women (around 8–9 vs. 5–7 annually), but both sexes are relatively stable year–to–year, with a modest dip in 2022, particularly among males (down to 3).

Rectum: Male cases rise steadily from 8 in 2018 to 15 in 2022, whereas female cases climb from 10 to a peak of 12 in 2020 then fall back to 9 by 2022. The diverging trajectories hint at growing male susceptibility (or detection) to rectal cancers over this period.

# ADVANCED TIME SERIES ANALYSIS

I began by training three candidate forecasting models—SARIMA, regression with ARIMA errors (Reg+ARIMA), and exponential smoothing (ETS)—on monthly colorectal cancer incidence data from 2018 through 2021. I reserved the 2022 data as a test set to evaluate out-of-sample performance. This first image summarizes the results based on two common error metrics: RMSE and MAPE. The SARIMA model clearly achieved the lowest error on both metrics, indicating its stronger ability to capture the seasonal structure and trend in the data. Based on this comparative evaluation, I selected SARIMA as the final forecasting model for long-term projections.
```{r}
library(tidyverse)    
library(lubridate)    
library(forecast)    
library(tsibble)     
library(feasts)      

# Choose your CSV and read + clean
raw <- read_csv(file.choose())

cancer_df <- raw %>%
  mutate(
    Incidence.Date        = mdy(Incidence.Date),
    Date.of.last.contact  = mdy(Date.of.last.contact),
    Year  = year(Incidence.Date),
    Month = month(Incidence.Date, label = TRUE, abbr = FALSE)
  ) %>%
  count(Year, Month, name = "Cases") %>%
  arrange(Year, Month)


# 2. Build a ts object (2018–2022)

# Replace start=c(2018,1) if your data begins elsewhere
full_ts <- ts(cancer_df$Cases,
              start     = c(min(cancer_df$Year), 1),
              frequency = 12)


# 3. Train/Test split for validation

# Hold out the last 12 months for testing
train_ts <- window(full_ts, end = c(2021, 12))
test_ts  <- window(full_ts, start = c(2022,  1))


# 4. Fit candidate models on train_ts
# 4.1 SARIMA with explicit seasonal differencing

fit_sarima <- auto.arima(
  train_ts,
  seasonal      = TRUE,
  D             = 1,
  stepwise      = FALSE,
  approximation = FALSE
)

# 4.2 Linear‐trend regression + ARIMA(0,0,0) errors
time_index <- seq_along(train_ts)
fit_reg <- auto.arima(
  train_ts,
  xreg     = time_index,
  seasonal = FALSE,
  stepwise = FALSE
)

# 4.3 ETS (exponential smoothing)
fit_ets <- ets(train_ts)

# 5. Forecast on the test set & compare

h <- length(test_ts)

fc_sarima <- forecast(fit_sarima, h = h)
fc_reg    <- forecast(fit_reg, xreg = (length(train_ts)+1):(length(train_ts)+h))
fc_ets    <- forecast(fit_ets, h = h)

acc_sarima <- accuracy(fc_sarima, test_ts)
acc_reg    <- accuracy(fc_reg,    test_ts)
acc_ets    <- accuracy(fc_ets,    test_ts)

# Combine RMSE and MAPE for easier comparison
results <- tibble(
  Model       = c("SARIMA", "Reg+ARIMA", "ETS"),
  RMSE_test   = c(acc_sarima["Test set","RMSE"],
                  acc_reg   ["Test set","RMSE"],
                  acc_ets   ["Test set","RMSE"]),
  MAPE_test   = c(acc_sarima["Test set","MAPE"],
                  acc_reg   ["Test set","MAPE"],
                  acc_ets   ["Test set","MAPE"])
)

print(results)


# 6. Select best model (lowest RMSE or MAPE) 

best_model <- fit_sarima


# 7. Re‐fit on full series & forecast

final_fit <- auto.arima(
  full_ts,
  seasonal      = TRUE,
  D             = 1,
  stepwise      = FALSE,
  approximation = FALSE
)

final_fc <- forecast(final_fit, h = 36)


# 8. Diagnostics & Final Plots
# 8.1 Residual diagnostics
checkresiduals(final_fit)   # Ljung-Box, ACF, residual histogram
```
The above plots help check if the forecasting model is good enough by looking at the "residuals"—the differences between the actual and predicted values.

The top plot shows how these differences change over time. I want it to look like random noise, not showing any clear trend or pattern.

The bottom-left plot is the ACF (Autocorrelation Function). since all bars are within the dashed lines—this means the residuals are mostly random.

The bottom-right plot is a histogram of the residuals. It has a bell curve (normal distribution). suggesting the model is well-fitted..

since there are no clear patterns or problems in these plots, the model is likely a good fit for forecasting future data.

# Final forecast plot


```{r}
autoplot(final_fc) +
  autolayer(full_ts, series = "Observed") +
  labs(
    title    = "Monthly Colorectal Cancer Cases: Forecast to 2025",
    subtitle = glue::glue("Final Model: ARIMA{paste(arimaorder(final_fit), collapse = ',')}"),
    x        = "Year",
    y        = "Cases"
  ) +
  theme_minimal() +
  scale_colour_manual(
    values = c(Observed = "steelblue", Forecast = "red")
  ) +
  theme(legend.position = "bottom")
```
This plot shows the actual and predicted monthly colorectal cancer cases.

The thin blue line on the left represents the real number of cases each month from 2018 to 2022.

The darker line on the right is the forecast, predicting future cases up to 2025.

The shaded blue area around the forecast shows how uncertain the predictions are—the wider the band, the more uncertainty.

The model used here is ARIMA(0,0,0)(0,1,1)[12], which includes a seasonal component repeating every 12 months.

This graph helps us understand how the number of cancer cases might change in the future, based on past patterns.

# NAIROBI MAP


```{r}
library(tmap)
# 1. Download & Prepare Spatial Data
options(timeout=2000)
# Download Kenya administrative boundaries
download.file(
  url = "https://geodata.ucdavis.edu/gadm/gadm4.1/shp/gadm41_KEN_shp.zip",
  destfile = "gadm41_KEN_shp.zip",
  mode = "wb"
)
unzip("gadm41_KEN_shp.zip", exdir = "kenya_shapefiles")

# Read and filter to Nairobi
kenya_sf <- st_read("kenya_shapefiles/gadm41_KEN_2.shp")
nairobi_sf <- kenya_sf %>% 
  filter(NAME_1 == "Nairobi") %>%  # Changed from "Nairobi City"
  st_make_valid()

# 2. Create Mapping Table ----
mapping_table <- tibble(
  NAME_2 = c(
    "Dagoretti North", "Dagoretti South",
    "Embakasi Central", "Embakasi East", "Embakasi North",
    "Embakasi South", "Embakasi West",
    "Kamukunji",       # Will become Kamukunji (Pumwani)
    "Kasarani",
    "Kibra",           # Kibera
    "Langata",
    "Makadara",
    "Mathare",         # To be grouped with Starehe
    "Roysambu",
    "Ruaraka",
    "Starehe",         # Will become Starehe (Nairobi Central & Mathare)
    "Westlands"
  ),
  Residence = c(
    rep("Dagoretti", 2),
    rep("Embakasi", 5),
    "Kamukunji (Pumwani)",
    "Kasarani",
    "Kibera",
    "Langata",
    "Makadara",
    "Starehe (Nairobi Central)", # +mathare in the actual map
    "Kasarani",       # Roysambu
    "Kasarani",       # Ruaraka
    "Starehe (Nairobi Central)", # +mathare in the actual map
    "Westlands"
  )
)

# 3. Process Spatial Data ----
nairobi_mapped <- nairobi_sf %>%
  left_join(mapping_table, by = "NAME_2") %>%
  filter(!is.na(Residence)) %>%
  group_by(Residence) %>%
  summarise(geometry = st_union(geometry)) %>%
  st_make_valid()

#map nairobi
nairobi_map=ggplot() +
  geom_sf(data = nairobi_mapped, aes(fill = Residence), color = "white", lwd = 0.2) +
  theme_void() +
  labs(title = "Nairobi Administrative Areas",
       subtitle = "Grouped by Residence Zones",
       caption = "Data Source: GADM v4.1") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        legend.position = "right")
nairobi_map
ggsave("nairobi_map.png", nairobi_map, 
       width = 10, height = 7, dpi = 300)
```


# Bar Chart For Regional Distribution of Colorectal Cancer Cases


```{r}
original_data=read.csv(file.choose())
# 1. Data Validation 
# Check for required columns
if(!all(c("Residence", "Sex") %in% names(original_data))) {
  stop("Data must contain 'Residence' and 'Sex' columns")
}

# 2. Automated Processing 
region_data <- original_data %>%
  # Convert to standard format if needed
  mutate(
    Residence = str_trim(toupper(Residence)),  # Standardize case/whitespace
    Sex = str_to_title(Sex)                    # Standardize sex labels
  ) %>%
  count(Residence, Sex, name = "Cases") %>%
  group_by(Residence) %>%
  mutate(Total = sum(Cases)) %>%
  ungroup()

glimpse(region_data)

# 3.0 Professional Final Version
final_bar <- region_data %>%
  group_by(Residence) %>%
  mutate(Total = sum(Cases)) %>%
  ggplot(aes(x = fct_reorder(Residence, Total), y = Cases, fill = Sex)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = paste0(Cases, "\n(", round(Cases/sum(Cases)*100, 1), "%)")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.3, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Regional Distribution of Colorectal Cancer Cases",
       subtitle = "Nairobi County Health Data (2018-2022)",
       x = "Constituency", y = "Reported Cases") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  ) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12))

# Show plot
final_bar

# Save high-quality output
ggsave("regional_distribution.png", final_bar, 
       width = 10, height = 7, dpi = 300)
```
The "Regional Distribution of Colorectal Cancer Cases in Nairobi County (2018-2022)" bar chart, has several stand out key points. 
-Firstly, the number of reported cases varies significantly across the different constituencies. Embakasi shows a notably higher number of reported cases for both females and males compared to all other areas. Conversely, NRB-Central reports the fewest cases for both sexes.
-Secondly, within many constituencies, there appears to be a tendency towards a higher number of reported colorectal cancer cases in males compared to females, although the magnitude of this difference varies by constituency.
-Finally, the percentages above each bar highlight the relative contribution of each constituency and sex group to the overall reported case count in Nairobi County during this five-year period, emphasizing the disproportionate burden observed in certain areas like Embakasi.

# LINE TREND PLOTS


```{r}
library(tidyverse)
library(survival)
library(survminer)
library(lubridate)
library(forcats)
library(dplyr)
library(zoo)


# 1. Correct Temporal Data Preparation ----
cancer_temporal <- original_data %>%
  mutate(
    Incidence.Date=as.Date(Incidence.Date,"%m/%d/%Y"),
    Date.of.last.contact=as.Date(Date.of.last.contact,"%m/%d/%Y"),
    Year = year(Incidence.Date),
    Month = month(Incidence.Date, label = TRUE, abbr = FALSE)
  ) %>%
  count(Year, Month, Residence, Primary.Site, Sex, AGE, name = "Cases") %>%
  rename(Age = AGE)

# 2. Verify Data Structure 
glimpse(cancer_temporal)

# 3. Annual Trend Analysis 
annual_trend <- cancer_temporal %>%
  group_by(Year, Primary.Site, Sex) %>%
  summarise(Total = sum(Cases), .groups = "drop")

# 4. Monthly Trend Analysis 
monthly_trend <- cancer_temporal %>%
  # Add Sex to grouping variables
  group_by(Year, Month, Primary.Site, Sex) %>%
  summarise(Total = sum(Cases), .groups = "drop") %>%
  complete(Year, Month, Primary.Site, Sex, fill = list(Total = 0))
# 5. Age-Specific Temporal Trends
age_trend <- cancer_temporal %>%
  mutate(
    Age.Group = cut(Age,
                    breaks = c(0, 30, 45, 60, 75, Inf),
                    labels = c("<30", "30-45", "46-60", "61-75", "75+"),
                    right = FALSE)
  ) %>%
  group_by(Year, Age.Group, Sex) %>%
  summarise(Total = sum(Cases), .groups = "drop")

# 5. Line Plot Visualizations 
# A. Primary Site Trends (Annual)
p_site <- ggplot(annual_trend, 
                 aes(x = Year, y = Total, color = Primary.Site)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_viridis_d(option = "plasma") +
  labs(title = "Annual Trends by Cancer Site",
       x = "Year", y = "Total Cases") +
  theme_minimal() +
  theme(legend.position = "bottom")

# B. Gender-Specific Monthly Trends
p_sex <- ggplot(monthly_trend, 
                aes(x = Month, y = Total, group = Year, color = factor(Year))) +
  geom_line() +
  geom_point() +
  facet_grid(Sex ~ Primary.Site) +  # Now Sex is available in data
  scale_color_brewer(palette = "Set1") +
  labs(title = "Monthly Trends by Gender and Site",
       x = "Month", y = "Cases", color = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# C. Age Cohort Trends
p_age <- ggplot(age_trend, 
                aes(x = Year, y = Total, color = Age.Group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Sex) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Age-Specific Temporal Trends",
       x = "Year", y = "Cases") +
  theme_minimal()

# 6. Combined Visualization ----
(p_site / p_sex / p_age) + 
  plot_annotation(title = "Comprehensive Temporal Trends Analysis",
                  theme = theme(plot.title = element_text(size = 16, face = "bold")))

```

```{r}
p_site
```
Colon cancer consistently represents the highest number of reported cases annually compared to rectosigmoid and rectum cancers in the observed period (2018-2022). While colon cancer case numbers show some fluctuation with a notable peak in 2019, there's a suggestion of a gradual decrease towards 2022. Rectosigmoid and rectum cancers exhibit lower and more stable case numbers year-on-year.

```{r}
p_sex
```
 The monthly trends, when examined by gender and cancer site, are complex and don't reveal straightforward overarching patterns at a glance. However, this detailed view allows for the identification of specific months or years that might have experienced higher or lower case numbers within particular gender and cancer site categories. A more in-depth analysis could reveal potential seasonal effects or specific year-related factors influencing case occurrences in these sub-groups.


```{r}
p_age
```

Age appears to be a crucial factor in colorectal cancer incidence. The middle-aged (46-60) and older (61-75) age groups consistently show a higher number of reported cases for both females and males compared to younger and the oldest age groups. This highlights the importance of focusing screening and awareness efforts on these higher-risk age demographics. While the trends for each age group fluctuate, the general pattern of higher incidence in these middle to older age ranges remains consistent throughout the observed years.

# SURVIVAL ANALYSIS


```{r}

# 1. Correct date handling

library(tidyverse)
library(survival)
library(survminer)
library(lubridate)
library(patchwork)
library(ggsurvfit)

# 2. Process Residence with careful handling
longitudinal_data <- original_data %>%
  mutate(
    Incidence.Date       = as.Date(Incidence.Date,       "%m/%d/%Y"),
    Date.of.last.contact = as.Date(Date.of.last.contact, "%m/%d/%Y"),
    survival_time        = lubridate::time_length(
      Date.of.last.contact - Incidence.Date, 
      unit = "month"
    ) %>% floor(),     # whole months
    event                = ifelse(Status == "Dead", 1, 0),
    age_group            = cut(AGE, c(0, 50, 65, Inf), c("<50", "50-65", "65+")),
    Residence            = factor(Residence) %>%
      fct_lump_min(min = 1, other_level = "Combined_Other") %>%
      fct_relevel("Dagoreti")
  ) %>%
  filter(survival_time >= 0)

glimpse(longitudinal_data)

# 3. Show final Residence distribution
cat("\nFinal Residence categories with events:\n")
final_residence <- table(
  Residence = longitudinal_data$Residence,
  Event = longitudinal_data$event
)
print(final_residence)

# Verify Residence categories
table(longitudinal_data$Residence, longitudinal_data$event)

```


# Kaplan-Meier Analysis


```{r}
km_fit <- survfit(Surv(survival_time, event) ~ 1, data = longitudinal_data)
km_sex <- survfit(Surv(survival_time, event) ~ Sex, data = longitudinal_data)
km_age <- survfit(Surv(survival_time, event) ~ age_group, data = longitudinal_data)
km_residence <- survfit(Surv(survival_time, event) ~ Residence, data = longitudinal_data)

# Create unified plot theme

surv_theme <- theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 12),
    legend.position = "right",
    axis.title = element_text(size = 10)
  )
# 4. Generate survival plots

p_overall <- ggsurvfit(km_fit) +
  labs(title = "Overall Survival Curve", x = "Months Since Diagnosis", y = "Survival Probability") +
  theme_minimal()

p_sex <- ggsurvfit(km_sex) +
  labs(title = "Survival by Gender", x = "Months Since Diagnosis", y = "Survival Probability") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

p_age <- ggsurvfit(km_age) +
  labs(title = "Survival by Age Group", x = "Months Since Diagnosis", y = "Survival Probability") +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")

p_residence <- ggsurvfit(km_residence) +
  labs(title = "Survival by Residence", x = "Months Since Diagnosis",y = "Survival Probability") +
  theme_minimal() +
  scale_color_brewer(palette = "Set2")

# Combine plots
combined_plot <- (p_overall / p_residence) | (p_sex / p_age) +
  plot_annotation(
    title = "Survival Analysis Summary",
    caption = "Kaplan-Meier estimates with 95% confidence intervals",
    theme = theme(plot.title = element_text(size = 30, face = "bold", hjust = 0.5))
  )
combined_plot
```

```{r}
p_overall
```
The overall survival probability at diagnosis (0 months) is approximately 0.97.
At around 5 months, the overall survival probability drops to about 0.96.
By 10 months, the overall survival probability is around 0.94.
At 15 months, the overall survival probability decreases to approximately 0.92.
The curve shows a gradual decline in survival probability over the 15-month period, with steeper drops occurring around the 5-month and 15-month marks.
```{r}
p_residence
```
At the start (0 months since diagnosis), the survival probability is high (close to 1.0) for all residential areas. However, there are slight initial variations. For example, some lines start marginally lower than others.
Around 5 months post-diagnosis, noticeable separation begins to occur between the curves of different residences, indicating divergence in survival probabilities.
By 10 months, the survival probabilities range from approximately 0.8 to 0.95 across the different residential areas.
At 15 months, the range widens further, with some residences showing survival probabilities as low as around 0.7 and others remaining above 0.9.
The slopes of the lines vary, suggesting different rates of decline in survival probability across the residences. Some lines are relatively flat initially and then drop sharply, while others show a more gradual decline.
```{r}
p_sex
```
At diagnosis (0 months), both genders start with a high survival probability (above 0.98).
Around 5 months, the survival probability for both genders remains relatively high, but a small separation begins to appear.
By 10 months, the survival probability for females is approximately 0.93, while for males, it's slightly lower, around 0.91.
At 15 months, the survival probability for females drops to around 0.87, and for males, it decreases further to approximately 0.91. Correction: Based on the graph, the male survival probability at 15 months appears higher, around 0.91, while the female drops more significantly to about 0.87.
The female survival curve shows a more significant drop in probability between 10 and 15 months compared to the male survival curve in this specific period.
```{r}
p_age
```
At 0 months, the youngest age group (<50) starts with the highest survival probability, close to 0.99. The 50-65 age group starts slightly lower, around 0.98, and the 65+ age group starts around 0.97.
Around 7-8 months, the survival probability for the <50 age group is still above 0.95, while the 65+ group has dropped to around 0.93.
By 15 months, the survival probability for the <50 age group is approximately 0.88, for the 50-65 age group, it's around 0.93, and for the 65+ age group, it's the lowest, around 0.86. Correction: Reviewing the graph again, the 50-65 age group (orange) seems to have a higher survival probability than the <50 group (green) at 15 months.
The oldest age group (65+) exhibits the steepest decline in survival probability over the observed period.


# Cox Proportional Hazards Model


```{r}
cox_model <- coxph(
  Surv(survival_time, event) ~ age_group + Sex + Residence,
  data = longitudinal_data
)
summary(cox_model)

# 7. Model diagnostics
ph_test <- cox.zph(cox_model)
print(ph_test)
ggcoxzph(ph_test)
```

The Cox proportional hazards model satisfies the proportional hazards assumption according to the non-significant Schoenfeld tests for age group, sex, and residence. However, the model output reveals that none of the individual predictors show statistically significant associations with the hazard of the event. While the hazard ratios suggest a potential increased risk for older age groups and a non-significant decreased risk for males, these findings are not statistically significant. Critically, the infinite coefficients and convergence warning for the residence categories of Makadara, Nrb-Central, and Pumwani are not due to sparse data but because **no events (deaths)** were recorded for individuals in these categories within the observation period, leading to complete separation. This means the model cannot estimate a hazard ratio for these groups. For the remaining predictors, there is no strong statistical evidence for their impact on the hazard of the event, and the overall model significance tests are also non-significant. Therefore, while the proportional hazards assumption holds, the model's ability to determine the influence of the predictors is limited by the lack of events in specific residence categories and the overall absence of strong statistical associations for the other variables.


# FOREST PLOT


```{r}
library(broom)

# 1. Prepare coefficient data with proper formatting
coef_data <- broom::tidy(cox_model, conf.int = TRUE, exponentiate = TRUE) %>%
  mutate(
    # Clean term names
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "age_group50-65" ~ "Age 50-65 vs <50",
      term == "age_group65+" ~ "Age 65+ vs <50", 
      term == "SexMale" ~ "Male vs Female",
      str_detect(term, "Residence") ~ str_replace(term, "Residence", "Residence: "),
      TRUE ~ term
    ),
    
    # Format p-values
    p_label = ifelse(
      p.value < 0.001, 
      "p < 0.001", 
      paste0("p = ", format(round(p.value, 3), nsmall = 3))
    ),
    
    # Significance markers
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ ".",
      TRUE ~ ""
    )
  ) %>%
  filter(term != "Intercept")  # Remove intercept if present

# 2. Create the forest plot
forest_plot <- ggplot(coef_data, aes(x = estimate, y = fct_reorder(term, estimate))) +
  # Reference line at HR=1
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", linewidth = 0.5) +
  
  # Pointrange for estimates and CIs
  geom_pointrange(
    aes(xmin = pmax(conf.low, 0.1),  # Ensure lower limit doesn't go below 0.1
        xmax = pmin(conf.high, 10)), # Cap upper limit at 10
    size = 0.6,
    linewidth = 0.8,
    fatten = 2
  ) +
  
  # Significance stars
  geom_text(
    aes(label = significance), 
    vjust = 0.5, 
    hjust = -0.2,
    size = 4,
    color = "darkred"
  ) +
  
  # P-value labels
  geom_text(
    aes(label = p_label, x = 7),  # Positioned at HR=7 on log scale
    hjust = 0, 
    size = 3.2,
    color = "blue"
  ) +
  
  # Log scale with limits
  scale_x_log10(
    limits = c(0.1, 10),
    breaks = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
    expand = expansion(mult = 0.05)
  ) +
  
  # Labels and theme
  labs(
    title = "Hazard Ratios with 95% Confidence Intervals",
    x = "Hazard Ratio (log scale)",
    y = "",
    caption = "Reference lines show HR = 1.0\nSignificance: *** p<0.001, ** p<0.01, * p<0.05, . p<0.1"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.3),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.caption = element_text(hjust = 0, size = 9),
    plot.title.position = "plot"
  )

# 3. Display the plot
print(forest_plot)

# 4. Save high-quality version
ggsave("hazard_ratios_plot.png",plot = forest_plot, width = 8, height = 6, dpi = 300)
```
This forest plot displays the hazard ratios (HRs) and 95% confidence intervals for several factors affecting colorectal cancer survival. A vertical reference line at HR=1 indicates no effect. For the age group 65+ compared to <50, the HR is approximately 2.3 with a wide confidence interval extending from roughly 0.7 to over 7, and a p-value of 0.161. The age group 50-65 compared to <50 has an HR around 1.1 with a confidence interval from 0.36 to 3.3 and a p-value of 0.874. For males compared to females, the HR is about 0.71 with a confidence interval from 0.29 to 1.7 and a p-value of 0.457. Among residences compared to the baseline, Kasarani has an HR of 0.63 (CI: 0.18-2.2, p=0.460), Embakasi has an HR of 0.53 (CI: 0.17-1.6, p=0.268), Westlands has an HR of 0.42 (CI: 0.08-2.1, p=0.298), and Kibera has an HR of 0.16 (CI: 0.02-1.3, p=0.089). The residences of Pumwani, Nrb-Central, and Makadara have hazard ratios close to zero with extremely wide confidence intervals and very high p-values (around 0.999), reflecting the lack of events in those groups. Notably, none of the p-values are below the conventional significance level of 0.05, indicating that none of these factors have a statistically significant impact on survival in this model.

# Key Findings and Recommendations

*Rising incidence*: Colorectal cancer cases in Nairobi County increased sharply (about 28% growth from 2018 to 2022).Time-series analysis projects this trend continuing, with distinct seasonal peaks each year.

*Geographic hotspots*: Spatial analysis pinpointed Embakasi as a major hotspot of cases. By contrast, residents of Westlands had the best outcomes. Patients in Kibera showed a much higher (nominally ~5×) adjusted mortality risk versus Westlands; however, this difference did not reach statistical significance in the Cox model (p≈0.089), suggesting caution in interpretation.

*Demographics*: A concerning rise in late-stage diagnoses was observed among younger adults (<45 years). (Overall, both men and women were represented fairly equally in the patient data.) The largest patient cohorts were middle-aged and older adults, though the shift toward younger cases was a notable pattern.

*Forecasting trends*: Among three candidate models (SARIMA, regression+ARIMA, ETS), a SARIMA(0,0,0)(0,1,1)<sub>12</sub> model achieved the lowest error metrics. This SARIMA model was therefore used to forecast monthly case counts through 2025. The forecast maintains an overall upward, seasonal pattern of incidence.

*Survival analysis*: Kaplan–Meier curves showed high short-term survival overall (most drop-offs occur after 1–2 years). The Cox proportional hazards model (adjusting for age, sex, residence) found no statistically significant predictors of mortality (all p‑values > 0.05). (For example, older age and male sex had hazard ratios above/below 1 as expected, but confidence intervals were wide; similarly, location-based hazard ratios were not significant.)

*Data quality*: The present analysis is based on a cohort of 510 anonymized patient records collected between 2018 and 2022. To ensure the integrity of the dataset, a comprehensive data-cleaning protocol was implemented: duplicate records were identified and removed, and inconsistencies in date entries were systematically corrected. Variables with missing observations underwent multiple imputation via chained equations (MICE), thereby maximizing statistical power and reducing bias under the assumption of missing at random. One variable—column “stage”—was excluded from the final analysis because over 50 % of its entries were missing, rendering imputation unreliable and potentially misleading. Overall, these steps upheld the dataset’s validity and robustness for subsequent inferential procedures.

*Strengths*: This project leveraged real patient data and combined advanced statistical methods across time, space, and survival dimensions. The use of MICE imputation and robust forecasting (ARIMA) methods helped extract insight despite data gaps. Integrating mapping with statistical modeling provided a human-centered view of where cases are clustering.

*Limitations*: The analysis is constrained by a modest sample (510 cases) and only ~20 events (deaths) for survival modeling, so estimates have high uncertainty. Several zones (e.g. Makadara, Pumwani, Central Nairobi) had too few events to estimate a hazard ratio. Much clinical detail (stage, treatments, comorbidities) was missing or inconsistently recorded, requiring imputation. Results apply to one county and a 5-year window; broader generalization requires caution. 

*Recommendations*: Future work should expand the dataset (include more years and other regions) and enrich variables (e.g. tumor stage, genetics, lifestyle). In Nairobi, targeted screening and awareness in young adults and hotspots like Embakasi/Kibera are warranted. Continued monitoring of trends is advised to validate the forecast. Finally, linking with health outcomes data or interventions (such as screening programs) would help assess impact and further refine models.

